Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-03-27 tagless/tagless-final.tex
Section 3 mostly done; added fodder for Sec 4

\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{natbib1}
\usepackage{comment}

@@

\begin{abstract}
We have built a (family of) tagless partial evaluator(s) for a typed object
language, placing minimal requirements on the type system of the metalanguage:
we need let-bound polymorphism and staging, but not dependent
types, generalized algebraic data types, nor a tag elimination 
postprocessing step.
The main tool is to take functions seriously: we encode the
higher-order abstract syntax of object terms not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of implicit representations of final algebras.
Our family of evaluators contains an interpreter, a compiler, a partial
evaluator and a term-size evaluator.
\end{abstract}

@@

The first attempt at the interpreter, the function eval. It takes the
object language term such as test1 and gives us its value.
The first argument to |eval| is the environment, initially empty,
which is the sequence of values associated with free variables in the
interpreted code.
\begin{code}
  let rec eval0 env = function 
  | V VZ -> List.hd env
  | V (VS v) -> eval0 (List.tl env) (V v)
  | B b -> b 
  | L e -> fun x -> eval0 (x::env) e
  | A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

@@

See the Related Work section (\ref{related}) for a discussion of the other
solutions to this problem.

@@

\jacques{do we want to start using a lot of little files and
use includes to insure the code stays in synch?}

@@

Thus all the errors that were previously reported at run-time (open terms,
applications of non-functions) are now reported at compile
time. Running of the interpreter will never result in run-time errors
because its code uses no operations that could possibly raise such
an error.

Also, note that 
the interpreter uses no Universal type nor pattern matching, and it
is obvious to the compiler that non-termination can only result from
interpreting |fix|.  If the source language is strongly normalizing, then
our interpreter is total.
\jacques{the language above has no fix (yet), so we should move this 
comment somewhere else}

@@

@@

We can also write a ``compiler'' from our lambda calculus directly
into OCaml code, using MetaOCaml's staging facilities.  This compiler
attempts no optimizations at all, it simply produces residual code.
This is a completely straightforward staging of |module R|.

% taken from incope.ml -- should this be put into another file, to make
% sure the paper does not get out of synch?
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
  let int (x:int) = .<x>.
  let bool (b:bool) = .<b>.
  let add e1 e2 = .<.~e1 + .~e2>.
  let mul e1 e2 = .<.~e1 * .~e2>.
  let leq x y = .< .~x <= .~y >.
  let eql x y = .< .~x = .~y >.
  let if_ eb et ee = 
    .<if .~eb then .~(et () ) else .~(ee () )>.

  let lam f = .<fun x -> .~(f .<x>.)>.
  let app e1 e2 = .<.~e1 .~e2>.
  let fix f = 
     .<let rec self n = .~(f .<self>.) n in self>.
  let dyn_id (x:('c,'sv,'dv) repr):('c,'sv1,'dv) repr = x
end;;
\end{code}

\noindent It is readily obvious that this compiler does not suffer
any interpretive overhead whatsoever, nor does the resulting code
contain any tags.  We add one item to the language, |dyn_id|, as it
is needed in the next section, where we will explain its purpose.

\jacques{we need to say something about the Haskell version here too,
even though we don't show it.  This is the first place we use GADT,
so maybe the ByteCode data type should be shown?}
\section{Tagless partial evaluator}\label{PE}

@@

  (* generic build - takes a repr constructor, an 
     interpreter function and a compiler function (all 
     binary) and builds a PE version *)
  let build cast f1 f2 = fun e1 -> fun e2 -> 
     match (e1,e2) with
         ({st = Some n1}, {st = Some n2}) -> cast (f1 n1 n2)
        | _ -> pdyn (f2 (abstr e1) (abstr e2))
  (* same as 'build' but takes care of the neutral element 
     (e) simplification allowed via a monoid structure which is 
     implicitly present *)
  let buildsimp cast e f1 f2 = fun e1 e2 -> 
     match (e1,e2) with
        | ({st = Some e'}, _) when e = e' -> e2
        | (_, {st = Some e'}) when e = e' -> e1
        | ({st = Some n1}, {st = Some n2}) -> cast (f1 n1 n2)
        | _ -> pdyn (f2 (abstr e1) (abstr e2))

  let add e1 e2 = buildsimp int 0 R.add C.add e1 e2
  let mul e1 e2 = buildsimp int 1 R.mul C.mul e1 e2
  let leq e1 e2 = build bool R.leq C.leq e1 e2
  let eql e1 e2 = build bool R.eql C.eql e1 e2
  let if_ eb et ee = match eb with
           {st = Some b} -> if b then et () else ee ()
         | _ -> pdyn (C.if_ (abstr eb) 
                          (fun () -> abstr (et ()))
                          (fun () -> abstr (ee ())))
  let lam f = {st = Some f; 
               dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with
        {st = Some f} -> f ea
      | _ -> pdyn (C.app (C.dyn_id (abstr ef)) (abstr ea))

   (* For now, to avoid divergence at the PE stage,
      we unroll the fixpoint exactly once, and then
      residualize *)
  let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
end;;
\end{code}

the sv type problem: the inductive type function 'dv -> 'sv
its GADT solution. 
its type-level function solution: incope1.hs

@@

Drawbacks of the GADT solution. GADTs, although work, are still not
quite satsifactory: let us look at the following lines in incope.hs
\begin{code}
    app ef ea = case ef of
                        VF f -> f ea
                        E  f -> E (app f (abstr ea))
\end{code}
The datatype has four constructors, |VI|, |VB|, |VF| and |E|. How come we used
only two of them in the above expression. It is obvious, one may say,
that the constructors |VI| and |VB| just can't occur, because of the
typing consideration. This is obvious to us -- but it is not obvious
to the compiler. The case matching above does look \emph{partial}. That
seems like a tenuous point -- but it is precisely the main point of
the whole tag elimination approach. In a typed tagged interpreter
there are lots of pattern-matching that looks partial but in reality
total. The ultimate goal is to make this totality syntactically
apparent.

@@

So we have to augment the full signature of Symantics (from that of
sec.~\ref{encoding}) to
\begin{code}
module type Symantics = sig
  type ('c,'sv,'dv) repr
  val int  : int  -> ('c,int,int) repr
  val bool : bool -> ('c,bool,bool) repr
  val add  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val mul  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val leq  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,bool,bool) repr
  val if_  : ('c,bool,bool) repr ->
             (unit -> ('c,'sa,'da) repr) ->
             (unit -> ('c,'sa,'da) repr) -> ('c,'sa,'da) repr 
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr)
    -> ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),'da->'db) repr
  val app : ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),'da->'db) repr
    -> ('c,'sa,'da) repr -> ('c,'sb,'db) repr
  val fix : (('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr) as 's,'da->'db) repr 
             -> ('c,'s,'da->'db) repr)  -> ('c,'s,'da->'db) repr
end
\end{code}

\jacques{We talk a lot about GADT, but show none.  I think we need to
show a bit of Haskell, to explain what we are talking about.}

@@

\subsection*{Acknowledgments}
We would like to thank Martin Sulzmann for helpful discussions.

>>>>

\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{natbib1}
\usepackage{comment}
\usepackage{url}

@@

\begin{abstract}
We have built a (family of) tagless partial evaluator(s) for a typed object
language \oleg{``,or, embedded DSL''}, placing minimal requirements on
the type system of the meta-language:
we need let-bound polymorphism and staging, but not dependent
types, generalized algebraic data types, nor a tag elimination 
postprocessing step.

The main tool is to take functions seriously: we encode the
higher-order abstract syntax of object terms not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of implicit representations of final algebras.
Our family of evaluators contains an interpreter, a compiler, a partial
evaluator and a term-size evaluator \oleg{and a CPS CBN transfromer?}
Our approach highlights the importance of order-2 type constructors
(order-2 polymorphism), which lets us encode DSL terms with
translucent types: enough abstraction is available to interpret the
terms in various ways, and yet enought type information is available 
to typecheck the terms and statically assure their interpretation will
never get stuck.
\oleg{say something about type functions, GADTs and the way to get
  around them?}
\end{abstract}

@@

The first attempt at the interpreter, the function eval. It takes the
object language term such as |test1| and gives us its value.
The first argument to |eval| is the environment, initially empty,
which is the sequence of values associated with free variables in the
interpreted code.
\begin{code}
  let rec eval0 env = function 
  | V VZ -> List.hd env
  | V (VS v) -> eval0 (List.tl env) (V v)
  | B b -> b 
  | L e -> fun x -> eval0 (x::env) e
  | A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

@@

See the Related Work section (\ref{related}) for a discussion of the other
solutions to this problem.
\oleg{I feel we need to give some details of the other solutions here, 
to give the reader the feeling of
how advanced solutions were proposed and how rarely they are
implemented.}

@@

\jacques{do we want to start using a lot of little files and
use includes to insure the code stays in synch?}
\oleg{Because the fragements are interrelated, we can't. We have to
  use sort of a `partial include', and I don't think that is easy. I
  have checked all the fragments so far.}

@@

Thus all the errors that were previously reported at run-time (open terms,
applications of non-functions) are now reported at compile
time. Running of the interpreter will never result in run-time errors
because its code uses no operations that could possibly raise such
an error. In fact, our interpreter is total.

@@

Also, note that the interpreter uses no Universal type nor pattern
matching, and it is obvious to the compiler that non-termination can
only result from interpreting |fix|.

@@

We can also write a ``compiler'' from our lambda calculus directly
into OCaml code, using MetaOCaml's staging facilities. Namely,
future-stage expressions of the type |t| are represented at the
present stage as values of the type |('a,t) code| where |'a| is the
environment classifier (Walid03, ESOP04). Code values are created by
a \emph{bracket} form: |.<e>.| specifies the expression |e| is to be
evaluated at the future stage. The \emph{escape} |.~e|, which must occur
within a bracket, specifies that the expression |e| must be evaluated
at the present stage; its result, which must be a code value, is
spliced in into the code being built. The \emph{run} form |.! e|
evaluates the code value, which involves run-time compilation and
linking of the future-stage code. The bracket, escape and run are akin to
quasi-quotation, unquotation and |eval| of Lisp.

% this code does not have the 'sv parameter. It shows up later.
\begin{code}
module C = struct
  type ('c,'dv) repr = ('c,'dv) code
  let int (x:int) = .<x>.
  let bool (b:bool) = .<b>.
  let add e1 e2 = .<.~e1 + .~e2>.
  let mul e1 e2 = .<.~e1 * .~e2>.
  let leq x y = .< .~x <= .~y >.
  let if_ eb et ee = 
    .<if .~eb then .~(et () ) else .~(ee () )>.

  let lam f = .<fun x -> .~(f .<x>.)>.
  let app e1 e2 = .<.~e1 .~e2>.
  let fix f = 
     .<let rec self n = .~(f .<self>.) n in self>.
end;;
\end{code}
This is a completely straightforward staging of
|module R|.
This compiler attempts no optimizations at all, it simply produces
residual code. For example, the evaluation of our |test1|
\begin{code}
  let module E = EX(C) in E.test1 ();;
\end{code}
gives the value of the inferred type |('a, bool) C.repr| which looks
like |.<((fun x_6 -> x_6) (true))>.|. MetaOCaml can print the code
values, that is, the corresponding future-stage code.

\begin{comment}
\oleg{one may wish to give the folloiwng example too; the code
  expression can be somewhat abbreviated...}
\begin{code}
  let module E = EX(C) in E.testpowfix7;;
# - : ('_a, int -> int) C.repr =
.<fun x_1 ->
   (((fun x_2 ->
       let rec self_3 =
        fun n_4 ->
         ((fun x_5 -> if (x_5 <= 0) then 1 else (x_2 * (self_3 (x_5 + (-1)))))
           n_4) in
       self_3) x_1) 7)>.
\end{code}
\end{comment}

\noindent It is readily obvious that this compiler does not suffer
any interpretive overhead whatsoever. Indeed, the
  code produced for a function is simply |fun x_6 -> x_6| and has no 
  embedded calls to the evaluator: cf. the evaluator of DSL functions
  in Section~\ref{tagproblem}, the L-line.
The resulting code obviously contains no tags and no pattern-matching.

We have also implemented the ``compiler'' in Haskell. Since Haskell
has no (convenient, and typed) staging facilities, we had to emulate
them. To be precise, we defined a data type |ByteCode| with the
constructors |Var|, |Lam|, |App|, |Fix|, |INT| etc. with the obvious
meaning. Unlike our staging language (which is based on higher-order
abstract syntax), our bytecode, to be realistic, uses integer-named
variables. We then define 
\begin{code}
  newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for the creation of fresh variable
names. We define the compiler by making |C| to be the instance of the
class |Symantics|. The implementation is quite similar (but more
verbose) than the corresponding MetaOCaml code given above. The
accompanying complete code gives more details.

\oleg{I wanted to avoid saying anything about GADTs here. There
  doesn't seem to be much point in getting into too long
  explanations...}

\section{Tagless partial evaluator}\label{PE}

@@

\subsection{Non-functorialness of functions problem}
the sv type problem: the inductive type function 'dv -> 'sv
its GADT solution. 
its type-level function solution: incope1.hs 
\begin{code}
data P t where
    VI :: Int -> P Int
    VB :: Bool -> P Bool
    VF :: (P a -> P b) -> P (a->b)
    E  :: C t -> P t

abstr :: P t -> C t
abstr (VI i) = int i
abstr (VB b) = bool b
abstr (VF f) = lam (abstr . f . E)
abstr (E x) = x

instance Symantics P where
    int x  = VI x
    bool b = VB b

    -- lam :: (repr a -> repr b) -> repr (a->b)
    lam = VF
    -- app :: repr (a->b) -> repr a -> repr b
    app ef ea = case ef of
                        VF f -> f ea
                        E  f -> E (app f (abstr ea))
    add e1 e2 = case (e1,e2) of
                 (VI n1,VI n2) -> VI nr where nr = n1 + n2
                 _ -> E (add (abstr e1) (abstr e2))
\end{code}

GADT: |E|, Purely dynamic; the other alternatives: |static| code.

@@

Drawbacks of the GADT solution. GADTs, although work, are still not
quite satisfactory: let us look at the following lines in incope.hs
\begin{code}
    app ef ea = case ef of
                        VF f -> f ea
                        E  f -> E (app f (abstr ea))
\end{code}
The datatype has four constructors, |VI|, |VB|, |VF| and |E|. How come we used
only two of them in the above expression. It is obvious, one may say,
that the constructors |VI| and |VB| just can't occur, because of the
typing consideration. This is obvious to us -- but it is not obvious
to the compiler. The case matching above does look \emph{partial}. That
seems like a tenuous point -- but it is precisely the main point of
the whole tag elimination approach. In a typed tagged interpreter
there are lots of pattern-matching that looks partial but in reality
total. The ultimate goal is to make this totality syntactically
apparent.

@@

So we have to augment the full signature of Symantics (from that of
sec.~\ref{encoding}) to
\begin{code}
module type Symantics = sig
  type ('c,'sv,'dv) repr
  val int  : int  -> ('c,int,int) repr
  val bool : bool -> ('c,bool,bool) repr
  val add  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val mul  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val leq  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,bool,bool) repr
  val if_  : ('c,bool,bool) repr ->
             (unit -> ('c,'sa,'da) repr) ->
             (unit -> ('c,'sa,'da) repr) -> 
             ('c,'sa,'da) repr 
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr)
    -> ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
           'da->'db) repr
  val app : ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
             'da->'db) repr
    -> ('c,'sa,'da) repr -> ('c,'sb,'db) repr
  val fix : 
    (('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr) as 's,
         'da->'db) repr 
    -> ('c,'s,'da->'db) repr)  -> ('c,'s,'da->'db) repr
end
\end{code}

%%  let dyn_id (x:('c,'sv,'dv) repr):('c,'sv1,'dv) repr = x

%% We add one item to the language, |dyn_id|, as it
%% is needed in the next section, where we will explain its purpose.

  (* generic build - takes a repr constructor, an 
     interpreter function and a compiler function (all 
     binary) and builds a PE version *)
  let build cast f1 f2 = fun e1 -> fun e2 -> 
     match (e1,e2) with
         ({st = Some n1}, {st = Some n2}) -> cast (f1 n1 n2)
        | _ -> pdyn (f2 (abstr e1) (abstr e2))
  (* same as 'build' but takes care of the neutral element 
     (e) simplification allowed via a monoid structure which is 
     implicitly present *)
  let buildsimp cast e f1 f2 = fun e1 e2 -> 
     match (e1,e2) with
        | ({st = Some e'}, _) when e = e' -> e2
        | (_, {st = Some e'}) when e = e' -> e1
        | ({st=Some n1}, {st=Some n2})-> cast (f1 n1 n2)
        | _ -> pdyn (f2 (abstr e1) (abstr e2))

  let add e1 e2 = buildsimp int 0 R.add C.add e1 e2
  let mul e1 e2 = buildsimp int 1 R.mul C.mul e1 e2
  let leq e1 e2 = 
     match (e1,e2) with
         ({st=Some n1},{st=Some n2})-> bool (R.leq n1 n2)
        | _ -> pdyn (C.leq (abstr e1) (abstr e2))
  let if_ eb et ee = match eb with
           {st = Some b} -> if b then et () else ee ()
         | _ -> pdyn (C.if_ (abstr eb) 
                          (fun () -> abstr (et ()))
                          (fun () -> abstr (ee ())))
  let lam f = {st = Some f; 
               dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with
        {st = Some f} -> f ea
      | _ -> pdyn (C.app (C.dyn_id (abstr ef)) (abstr ea))

  let fix f = 
    let fdyn = C.fix (fun x -> abstr (f (pdyn x))) in
    let fdynn e = pdyn (C.app fdyn e.dy) in
    {st=Some (function {st = Some _} as e -> 
               let rec self n = 
                 (match n.st with Some _-> app (f (lam self)) n 
                                 | _ -> fdynn n) in 
               app (f (lam self)) e
             | e  -> fdynn e);
    dy = fdyn }
end;;
\end{code}

@@

separate compilation: separate
typechecking of terms without knowing which particular
hole they will be plugged into. Formulate this as a prop?

\subsection*{Acknowledgments}
We would like to thank Martin Sulzmann and Walid Taha 
for helpful discussions.
